---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SGLRT

<!-- badges: start -->
<!-- badges: end -->

`SGLRT` is a R package implementation of Sequential Generalized Likelihood Ratio (GLR)-like Tests and confidence sequences in 

__Nonparametric iterated-logarithm extensions to Lorden's treatment of the sequential GLRT__  
J. Shin, A. Ramdas, A. Rinaldo  [arXiv](https://arxiv.org/abs/2010.08082)   


## Installation

You can install the `SGLRT` package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("shinjaehyeok/SGLRT_paper")
```

<!---You can install the released version of SGLRT from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("SGLRT")
```
-->

## How to reproduce all plots and simulation results in SRR'20.

To reproduce plots, you need to install `latex2exp` package which parses and converts LaTeX math formulas to R’s plotmath expressions. It is is not installed, you can run the following command to install is.
``` r
install.packages("latex2exp")
```

### 1. Compare boundaries for sequential GLR tests. *(Fig. 3 in Section III-A)*

The following Rcode reproduces the Fig. 3 in Section III-A in which we compared Lorden's and ours boundary values of sequential GLR test for normal distributions.

```{r Fig.3}
library(latex2exp)
library(SGLRT)


alpha <- 10^seq(-1,-10)
theta_vec <- c(1,0.5, 10^seq(-1,-10)) 
theta_vec <- 2^seq(0,-30)

d_vec <- theta_vec^2 / 2

for (i in seq_along(alpha[1:3])){
  f_lorden <- function(d) const_boundary_lorden(alpha[i], d)
  lorden <- sapply(d_vec, f_lorden)
  f_ours <- function(d) const_boundary(alpha[i], d)
  ours  <- sapply(d_vec, f_ours)
  if (i == 1){
    plot(1/theta_vec, unlist(lorden["g",]), type = "l", 
         log ="x",
         ylab = "Boundary Value",
         xlab = TeX("$|\\mu_1 - \\mu_0|^{-1}$ (log scale)"))
    points(1/theta_vec, unlist(ours["g",]), type = "l", col = 2)
  } else {
    points(1/theta_vec, unlist(lorden["g",]), type = "l", col = 1,
           lty = i)
    points(1/theta_vec, unlist(ours["g",]), type = "l", col = 2,
           lty = i)
  }
}
legend("topleft", 
       TeX(c(paste0("Lorden's ($\\alpha = ", alpha[1:3],"$)"), 
             paste0("Ours ($\\alpha = ", alpha[1:3],"$)"))),
       col = c(rep(1,3), rep(2,3)), lty = rep(1:3,2))
```

### 2. Ratio of CI's width to CLT *(Fig. 5 in Section IV-C)*

The following Rcode reproduces the Fig. 5 in Section IV-C in which we compared ratios of widths of confidence intervals to the pointwise and asymptotically valid normal confidence intervals based on the central limit theorem. To be specific, `Chernoff` is the nonasymptotic but pointwise valid confidence interval based on the Chernoff bound. `Stitching (HRMS'20)` and `Normal Mix. (HRMS'20)` are nonasymptotic and anytime-valid confidence intervals presented by [Howard et al., (2020+)](https://arxiv.org/abs/1810.08240).  `GLR-like  (Ours)` and `Discrete Mix. (Ours)` are two nonasymptotic and anytime-valid confidence intervals proposed in our paper.   `GLR-like` confidence intervals are based on the GLR statistics and its nonparametric extension, called GLR-like statistics. `GLR-like` confidence intervals are time-uniformly close to the Chrenoff bound on any given target time interval. `Discrete Mix.` confidence intervals are refined version of `GLR-like` ones. From the construction, `Discrete Mix.` confidence intervals are always tighter than their corresponding `GLR-like` confidence intervals. In the following plots, we use two different target time intervals to build `GLR-like` and `Discrete Mix.` confidence intervals. 


```{r Fig.5}
library(SGLRT)

# Chernoff
chornoff <- function(v, alpha = 0.025){
  sqrt(2  * log(1/alpha) / v)
}

# Normal mixture function
normal_mix <- function(v, alpha = 0.025, rho = 1260){
  sqrt(2 * (1/v + rho / v^2) * log(1/(2*alpha) * sqrt((v + rho)/rho) + 1)) 
}

#Stitching
stitch <- function(v, alpha = 0.025){
  1.7 * sqrt((log(log(2) + log(v)) + 0.72 * log(5.2/alpha)) / v) 
}

# CLT
CLT <- function(v, alpha = 0.025){
  qnorm(1-alpha) / sqrt(v)
}

# Functions to construct our bounds
# First one has target interval [1, nmax]
# Second one has target interval [nmax / 20, nmax * 4]

alpha <- 0.025
nmax <- 1e+5

nmax1 <- nmax
nmin1 = 1

nmax2 <- nmax1 * 4
nmin2 <- nmax1 / 20


ours1 <- SGLR_CI_additive(alpha, nmax1, nmin1)
ours2 <- SGLR_CI_additive(alpha, nmax2, nmin2)


# Compute the width of CIs on exponentially-spaced grid.
M <- log(nmax, base = 10)
v <- 10^(seq(0,M + 0.2, length.out = 100))


# Compute existing bounds
chornoff_vec <- sapply(v, chornoff)
CLT_vec <- sapply(v, CLT)
normal_mix_vec <- sapply(v, normal_mix)
stit_vec <- sapply(v, stitch)

existing_list <- list(stitch = stit_vec,
                      normal_mix = normal_mix_vec)

# Compute our bounds
GLR_like_1_vec <- sapply(v, ours1$GLR_like_fn)
GLR_like_2_vec <- sapply(v, ours2$GLR_like_fn)
our_dis_mix_1_vec <- sapply(v, ours1$dis_mix_fn)
our_dis_mix_2_vec <- sapply(v, ours2$dis_mix_fn)


ours_list_1 <- list(GLR_like_1 = GLR_like_1_vec,
                    our_dis_mix_1 = our_dis_mix_1_vec)

ours_list_2 <- list(GLR_like_2 = GLR_like_2_vec,
                    our_dis_mix_2 = our_dis_mix_2_vec)

# Plot ratio of bounds
title <- "Ratio of CI's width to CLT"
plot(v, chornoff_vec / CLT_vec, type = "l",
     main = title,
     ylab = "Ratio",
     xlab = "n",
     ylim = c(1, 4),
     xlim = c(1, nmax))
col = 1
legend_col <- c(1)
for (i in seq_along(existing_list)){
  col <- col + 1
  legend_col <- c(legend_col, col)
  lines(v, existing_list[[i]] / CLT_vec, col = col)
}
legend_lty <- rep(1, length(existing_list) + 1)
for (i in seq_along(ours_list_1)){
  col <- col + 1
  legend_col <- c(legend_col, col)
  lines(v, ours_list_1[[i]] / CLT_vec,
        lty = 2, lwd = 2, col = col)
}
legend_lty <- c(legend_lty, rep(2, length(ours_list_1)))
for (i in seq_along(ours_list_2)){
  col <- col + 1
  legend_col <- c(legend_col, col)
  lines(v, ours_list_2[[i]] / CLT_vec,
        lty = 4, lwd = 2, col = col)
}
legend_lty <- c(legend_lty, rep(4, length(ours_list_2)))
abline(v = c(nmin1, nmin2, nmax1, nmax2), lty = 3)
bounds_name <- c("Chernoff", "Stitching (HRMS'20)", "Normal Mix. (HRMS'20)",
                 "GLR-like 1 (Ours)", "Discrete Mix. 1 (Ours)",
                 "GLR-like 2 (Ours)", "Discrete Mix. 2 (Ours)")
legend("topright", bounds_name,
       lty = legend_lty,
       col = legend_col,
       bg= "white")
```

### 3. Efficiency of GLR-like and discrete mixture tests *(Table I-VI in Appendix D)*

The following Rcode reproduces Table I-VI in Appendix D in which we compared performances of sequential `GLR-like` and `Discrete mixture` tests to standard fixed sample size based tests for Gaussian and Bernoulli observations.  In this document, we repeated the simulation only `100` times to save the computation time but, in the paper, the simulation result based on `2000` times repetition is presented. 

```{r simulation functions}
# Type 1 and Type 2 error control simulation
library(SGLRT)

# Define functions for tests based on a fixed sample size.

# Compute power of Z-test
G_pwr <- function(n, mu_target, mu_0, alpha){
  thres_exact <- qnorm(alpha, mu_0, sd = 1/sqrt(n), lower.tail = FALSE)
  pwr <- pnorm(thres_exact, mu_target, sd = 1/sqrt(n), lower.tail = FALSE)
  return(pwr)
}

# Z-test function (reject the null if the returned value is positive)
z_test <- function(x_bar, n, mu_0, alpha){
  z_alpha <- qnorm(1-alpha)
  x_bar - mu_0 - z_alpha / sqrt(n)
}

# Compute power of binomial test
binom_pwr <- function(n, mu_target, mu_0, alpha){
  n <- floor(n)
  thres_exact <- qbinom(alpha, n, mu_0, lower.tail = FALSE)
  pwr <- pbinom(thres_exact, n, mu_target, lower.tail = FALSE)
  return(pwr)
}

# Exact binomial test (reject the null if the returned value is positive)
binom_test <- function(x_bar, n, mu_0, alpha){
  thres_exact <- qbinom(alpha, n, mu_0, lower.tail = FALSE)
  return(x_bar * n - thres_exact)
}

# Sequential tests

# GLR-like and discrete mixture test for sub-Gaussian
seq_G_test_generator <- function(alpha, nmax, nmin){
  G_out <- SGLR_CI(alpha, nmax, nmin)
  return(G_out)
} 

# GLR-like and discrete mixture test for sub-Bernoulli
ber_fn_list <- generate_sub_ber_fn()
seq_ber_test_generator <- function(alpha, nmax, nmin){
  ber_out <- SGLR_CI(alpha,
                     nmax, 
                     nmin,
                     breg = ber_fn_list$breg,
                     breg_pos_inv = ber_fn_list$breg_pos_inv,
                     breg_neg_inv = ber_fn_list$breg_neg_inv,
                     breg_derv = ber_fn_list$breg_derv,
                     mu_lower = ber_fn_list$mu_lower,
                     mu_upper = ber_fn_list$mu_upper,
                     grid_by = ber_fn_list$grid_by)
  return(ber_out)
} 

# Sample generators
# Gaussian samples
G_sample <- function(n, mu_true){
  rnorm(n, mean = mu_true)
}

# Bernoulli samples
ber_sample <- function(n, mu_true){
  rbinom(n, 1, prob = mu_true)
}

# Function to summarize the simulation result
summ_simul <- function(result_name,
                       simul_out,
                       mu_true_vec){
  out <-   data.frame(mu_true = mu_true_vec, 
                      exact_hacking = numeric(length(mu_true_vec)),
                      GLR_like = numeric(length(mu_true_vec)),
                      dis_mix = numeric(length(mu_true_vec)),
                      exact_test = numeric(length(mu_true_vec)))
  for (i in seq_along(mu_true_vec)){
    out[i, -1] <- simul_out[[i]][[result_name]]
  }
  return(out)
}
```

#### 3-1. Gaussian case
In the Gaussian setting, we compared sequential `GLL-like` and `Discrete mixture` tests to the `Z-test`. We set the null hypothesis as the mean of the Gaussian distribution is less than or equal to `0`, and the alternative hypothesis as the true mean is greater than `0.1`. The `Z-test` was performed based on a fixed sample size to make the test control both type-1 and type-2 errors by `0.1`. In the following three tables, results of all three testing procedures are summarized. In the first table, we show that, for each underlying true mean, how frequently each testing procedure rejects the null hypothesis. Here the column `Z-test (p-hacking)` represents the naive usage of `Z-test` as a sequential procedure in which we stop and reject the null whenever p-value of the `Z-test` goes below the level `0.1`. This is an example of p-hacking which inflates the type-1 error significantly larger than the target level. From the first table, we can check that the `Z-test` with continuous monitoring of p-values yields a large type 1 error (`0.43`) even under a null distribution (`mean = -0.5`) which is a way from the boundary of the null.  In contrast, for all other testing procedures, type 1 errors are controlled under the null distributions. 

For alternative distributions (`mean > 0.1`), the `Z-test` has larger powers than the prespecified bound (`0.9`) as expected. Note that the `discrete mixture` based sequential test achieves almost the same power compared to the `Z-test`. However, as we can check from the second and third tables, under the alternative distributions, the `discrete mixture` test detects the signal faster than the `Z-test` both on average and with high probability. The `GLR-like` test yields a weaker power at the boundary of the alternative space but it achieves higher powers and smaller sample size as the underlying true means being farther away from the boundary. 

However, `GLR-like` and `discrete mixture` tests do not always perform better than the `Z-test`. If the underlying true mean lies between boundaries of null and alternative spaces, both test have weaker power and requires more samples to detect the signal compared to the `Z-test`. Therefore, we recommend to set the boundary of the alternative conservatively in practice. 

```{r tables_G, cache = TRUE}
# Gaussian simulation
set.seed(1)
f_simul <- function(mu) {
  out <- test_simul(mu_target = 0.1,
                    mu_true = mu,
                    mu_0 = 0,
                    alpha = 0.1,
                    beta = 0.1,
                    B = 100,
                    print_progress = FALSE,
                    print_result = FALSE,
                    sample_generator = G_sample,
                    power_fn = G_pwr,
                    fixed_test_fn = z_test,
                    seq_test_fn = seq_G_test_generator)
  return(out)
}

mu_true_vec <- seq(-0.05, 0.2, by = 0.05)
simul_G_out <- lapply(mu_true_vec, f_simul)

reject_rate_G <- summ_simul("reject_rate", 
                          simul_G_out,
                          mu_true_vec)
sample_size_G <- summ_simul("sample_size",
                          simul_G_out,
                          mu_true_vec)
early_stop_ratio_G <- summ_simul("early_stop_ratio",
                               simul_G_out,
                               mu_true_vec)

```

```{r echo = FALSE, results = 'asis'}
print_type <- "html"
print(xtable::xtable(reject_rate_G,
                     caption = "Estimated probabilities of rejecting the null hypothesis. (Gaussian)"),
      type = print_type)

print(xtable::xtable(sample_size_G[c(1,3,4,5)],
                     caption = "Estimated average sample sizes of testing procedures. (Gaussian)"),
      type = print_type)
print(xtable::xtable(early_stop_ratio_G[c(1,3,4)],
                     caption = "Estimated probabilities of tests being stopped earlier than Z-test."),
      type = print_type)
```

#### 3-2. Bernoulli case
In the Bernoulli setting, we compared sequential `GLL-like` and `Discrete mixture` tests to the `exact binomial` test. We set the null hypothesis as the mean of the Bernoulli distribution is less than or equal to `0.1`, and the alternative hypothesis as the true mean is greater than `0.12`. The `exact binomial` test was performed based on a fixed sample size to make the test control both type-1 and type-2 errors by `0.1`. The following three tables summarize the simulation result. In all three tables, we can check the same pattern we observed from the Gaussian case. In the first table, we show that, for each underlying true mean, how frequently each testing procedure rejects the null hypothesis. Here the column `Exact binomial (p-hacking)` represents the naive usage of the exact test as a sequential procedure in which we stop and reject the null whenever p-value of the test goes below the level `0.1`. As we observed before, the p-hacking inflates the type-1 error significantly larger than the target level. From the table, we can check that the `exact binomial test` with continuous monitoring of p-values yields a large type 1 error (`0.23`) even under a null distribution (`mean = 0.09`) which is a way from the boundary of the null.  In contrast, for all other testing procedures, type 1 errors are controlled under the null distributions. 

For alternative distributions (`mean > 0.12`), the `exact binomial` test has larger powers than the prespecified bound `0.9` as expected. Again, as same as the Gaussian case, the `discrete mixture` based sequential test achieves almost the same power compared to the `exact binomial` test. However, as we can check from the second and third tables, under the alternative distributions, the `discrete mixture` test uses, on average and with a high probability, smaller numbers of samples to detect the signal than the `exact binomial` test with a fixed sample size. The `GLR-like` test yields a weaker power at the boundary of the alternative space but it achieves higher powers and smaller sample size as the underlying true means being farther away from the boundary. 

However, as same as the Gaussian case, `GLR-like` and `discrete mixture` tests do not always perform better than the `exact binomial` test. If the underlying true mean lies between boundaries of null and alternative spaces, both test have weaker power and requires more samples to detect the signal compared to the `exact binomial`. Therefore, it is recommended to set the boundary of the alternative close to the boundary of the null in practice. 



```{r tables_ber, cache = TRUE}
# Bernoulli simulation
set.seed(1)
f_simul <- function(mu) {
  out <- test_simul(mu_target = 0.12,
                    mu_true = mu,
                    mu_0 = 0.1,
                    alpha = 0.1,
                    beta = 0.1,
                    B = 100,
                    print_progress = FALSE,
                    print_result = FALSE,
                    sample_generator = ber_sample,
                    power_fn = binom_pwr,
                    fixed_test_fn = binom_test,
                    seq_test_fn = seq_ber_test_generator)
  return(out)
}

mu_true_vec <- seq(0.09, 0.14, by = 0.01)
simul_ber_out <- lapply(mu_true_vec, f_simul)

reject_rate_ber <- summ_simul("reject_rate", 
                          simul_ber_out,
                          mu_true_vec)
sample_size_ber <- summ_simul("sample_size",
                          simul_ber_out,
                          mu_true_vec)
early_stop_ratio_ber <- summ_simul("early_stop_ratio",
                               simul_ber_out,
                               mu_true_vec)
```

```{r echo = FALSE, results = 'asis'}
print_type <- "html"
print(xtable::xtable(reject_rate_ber,
                     caption = "Estimated probabilities of rejecting the null hypothesis. (Bernoulli)"),
      type = print_type)

print(xtable::xtable(sample_size_ber[c(1,3,4,5)],
                     caption = "Estimated average sample sizes of testing procedures. (Bernoulli)"),
      type = print_type)
print(xtable::xtable(early_stop_ratio_ber[c(1,3,4)],
                     caption = "Estimated probabilities of tests being stopped earlier than the exact binomial test."),
      type = print_type)
```
